Shallow EDSLs and Object-Oriented Programming

### 2016-06-28 paper/sections/Interpretations.tex
Revision

A well-known limitation of shallow embeddings is that they allow only a single
interpretation. Gibbons and Wu worked around this problem by accommodating
multiple interpretations using tuples. However, their encoding needs to modify
the original code, and thus is non-modular. This section illustrates how various type of
interpretations can be \emph{modularly} defined in OOP.
\begin{comment}
Although a modular solution based on \cite{swierstra2008data}
is also presented, it complicates the encoding dramatically and may prevent pratical use.
OO approach, on the contrary, provides modular yet simple solution of defining
multiple interpretations. 
\end{comment}

\subsection{Multiple Interpretations}\label{subsec:multiple}
%Defining additional interpretations is not trivial for shallow embedding,
%especially for functional languages.
\paragraph{Multiple Interpretations in Haskell}
Suppose that we want to have an additional function that checks whether a circuit is
constructed correctly. Gibbons and Wu's solution is:
\lstinputlisting[linerange=2-9]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
\noindent This solution is not modular because it relies 
on defining the two interpretations (\lstinline{width} and
\lstinline{wellSized}) simultaneously, using a pair. It is not
possible to define \lstinline{width} independently and then reuse it.
Whenever a new interpretation is needed (e.g. \lstinline{wellSized}), the
original code has to be revised -
the arity of the tuple must be incremented and the new interpretation has to be
appended to each case.
%we add the definition of \lstinline{wellSized} by modifying the original code.

\paragraph{Multiple Interpretations in Scala}
In contrast, new interpretations are introduced in a 
modular way with an OO language like Scala, as shown in Figure~\ref{code:operation}.
\begin{figure}
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
\caption{Adding new interpretations}
\label{code:operation}
\end{figure}
%Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding relies on three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
complementing the body of \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit} type
to avoid type mismatches in methods~\cite{eptrivially16}.

@@

\subsection{Dependent Interpretations}
 \emph{Dependent interpretations} are a generalization of multiple
interpretations, where one or more interpretations can use other
interpretations in their definition. Typically such interpretations 
would be said to be \emph{non-compositional} since they do not 
dependent only on themselves. 
In Haskell dependent interpretations are once again defined with
tuples in a non-modular way. We omit the Haskell code here, since 
it is quite similar to the example in Section~\ref{subsec:multiple}

@@

\subsection{Discussion} 
Gibbons and Wu~\cite{gibbons2014folding} claim that in shallow
embeddings new language constructs are easy to add, but new
interpretations are hard. As our OOP approach shows, in OOP both
language constructs and new interpretations are easy to add in shallow
embeddings. In other words, the circuit DSL presented so far does not
suffer from the Expression Problem. The key point is that procedural
abstraction combined with OOP features (subtyping, inheritance and
type-refinement) adds expressiveness over traditional procedural
abstraction. Gibbons and Wu do discuss a number of advanced techniques that 
can solve some of the modularity problems. For example, using type
classes, the \emph{finally
  tagless}~\cite{carette2009finally} can deal with the example in
Section~\ref{subsec:multiple}. However
tuples are still needed 
to deal with dependent interpretations. In contrast the approach
proposed here is just straightforward OOP, and dependent
interpretations are not a problem.
\begin{comment}
and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type. However, it forces dependent interpretations to be defined along with what they depend on.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
\end{comment}
Gibbons and Wu also show some different variants of interpretations,
such as context-sensitive interpretations. 
Context-sensitive transformations are unproblematic as well. 
For space reasons, full details are only available online.

>>>>

A often stated limitation of shallow embeddings is that they allow only a single
interpretation. Gibbons and Wu worked around this problem by using tuples. However, their encoding needs to modify
the original code, and thus is non-modular. This section illustrates how various type of
interpretations can be \emph{modularly} defined in OOP.
\begin{comment}
Although a modular solution based on \cite{swierstra2008data}
is also presented, it complicates the encoding dramatically and may prevent pratical use.
OO approach, on the contrary, provides modular yet simple solution of defining
multiple interpretations. 
\end{comment}

\subsection{Multiple Interpretations}\label{subsec:multiple}
%Defining additional interpretations is not trivial for shallow embedding,
%especially for functional languages.
\paragraph{Multiple Interpretations in Haskell}
Suppose that we want to have an additional function that checks whether a circuit is
constructed correctly. Gibbons and Wu's solution is:
\lstinputlisting[linerange=2-9]{./code/NewSemantics.hs}%APPLY:SEMANTICS_HS
\noindent This solution is not modular because it relies 
on defining the two interpretations (\lstinline{width} and
\lstinline{wellSized}) simultaneously, using a tuple. It is not
possible reuse the independently defined \lstinline{width} function in
Section~\ref{subsec:shallow}.
Whenever a new interpretation is needed (e.g. \lstinline{wellSized}), the
original code has to be revised:
the arity of the tuple must be incremented and the new interpretation has to be
appended to each case.
%we add the definition of \lstinline{wellSized} by modifying the original code.

\paragraph{Multiple Interpretations in Scala}
In contrast, Scala allows new interpretations to be introduced in a 
modular way, as shown in Figure~\ref{code:operation}.
\begin{figure}
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
\caption{Adding new interpretations}
\label{code:operation}
\end{figure}
%Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding relies on three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
implementing \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit} type
to avoid type mismatches in methods~\cite{eptrivially16}.

@@

\subsection{Dependent Interpretations}
 \emph{Dependent interpretations} are a generalization of multiple
interpretations, where one or more interpretations can use other
interpretations in their definition. Typically such interpretations 
would be said to be \emph{non-compositional} since they do not 
depend only on themselves. 
In Haskell dependent interpretations are once again defined with
tuples in a non-modular way. We omit the Haskell code here, since 
it is quite similar to the example in Section~\ref{subsec:multiple}

@@

\subsection{Discussion} 
Gibbons and Wu~\cite{gibbons2014folding} claim that in shallow
embeddings new language constructs are easy to add, but new
interpretations are hard. As our OOP approach shows, in OOP both
language constructs and new interpretations are easy to add in shallow
embeddings. In other words, the circuit DSL presented so far does not
suffer from the Expression Problem. The key point is that procedural
abstraction combined with OOP features (subtyping, inheritance and
type-refinement) adds expressiveness over traditional procedural
abstraction. Gibbons and Wu do discuss a number of advanced techniques that 
can solve some of the modularity problems. For example, using type
classes, \emph{finally
  tagless}~\cite{carette2009finally} can deal with the example in
Section~\ref{subsec:multiple}. However
tuples are still needed 
to deal with dependent interpretations. In contrast the approach
proposed here is just straightforward OOP, and dependent
interpretations are not a problem.
\begin{comment}
and \emph{data types a la
  carte}~\cite{swierstra2008data} (DTC).
Finally tagless approach uses a type class to abstract over all interpretations
of the language. Concrete interpretations are given through creating a data type and
making it an instance of that type. However, it forces dependent interpretations to be defined along with what they depend on.
DTC represents language constructs separately and composes them together using
extensible sums. However, not like OO languages which come with subtyping, one
has to manually implement the subtyping machinery for variants.
\end{comment}
Gibbons and Wu also show some different variants of interpretations,
such as context-sensitive interpretations. 
Context-sensitive transformations are unproblematic as well. 
For space reasons, full details are only available online.

### 2016-06-28 paper/sections/Introduction.tex
Revision

Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. The {\sc Composite} or {\sc Interpreter}
patterns are normally accepted to provide a way to encode ASTs. Thus, 
one possible interpretation is that \emph{according to definition of deep embedding
  above, the OO approach classifies as a deep
  embedding}.

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. 
%This should
%not came as a surprise, as OOP languages have evolved over more than 50
%years to improve the use of procedural abstraction. 
In order to understand how OOP can help with shallow embeddings, lets
first review some of the limitations commonly
found in the literature:

\begin{enumerate}

@@

\end{enumerate}

@@

We show that OOP abstractions, including \emph{(multiple)
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. Although from the
\emph{syntactical} point of view there are minor inconveniences in
the Scala version, from the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages.

>>>>

\noindent Although the above definition is quite reasonable and widely accepted,
it leaves some space to (mis)interpretation. For example it is unclear 
how to classify an EDSL implemented using the {\sc Composite} or {\sc Interpreter} 
patterns in OOP. Would this OO approach be
classified as a shallow or deep embedding? We believe arguments can be
made both ways. 
%The {\sc Composite} or {\sc Interpreter}
%patterns are normally accepted to provide a way to encode ASTs. Thus, 
%one possible interpretation is that \emph{according to definition of deep embedding
%  above, the OO approach classifies as a deep
%  embedding}.

@@

The main goal of this paper is to argue that OOP languages have
advantages for the implementation of shallow embeddings. 
%This should
%not came as a surprise, as OOP languages have evolved over more than 50
%years to improve the use of procedural abstraction. 
In order to understand how OOP can help with shallow embeddings, it is
helpful to
first review some of the limitations commonly
found in the literature:
\vspace{-5pt}
\begin{enumerate}

@@

\end{enumerate}
\vspace{-5pt}

@@

We show that OOP abstractions, including \emph{
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanims enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

We make our arguments by taking a recent paper by Gibbons and Wu~\cite{gibbons2014folding},
where procedural abstraction is used in Haskell to model a simple
EDSL, and we recode that EDSL in Scala. From the \emph{semantic} and \emph{modularity} point of view the
Scala version has clear advantages over the Haskell version.

### 2016-06-28 paper/sections/ShallowOO.tex
Revision

The DSL consists of three constructs - two primitives
\emph{id} and \emph{fan} and one combinator \emph{beside}.
Their meanings are: \emph{id n} contains \emph{n} isolated vertical wires;
\emph{fan n} has \emph{n} vertical wires with its first wire connected to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
With these three constructs, simple circuits can be described.
%For example, Figure~\ref{} visualize the circuit \emph{beside (fan 3) (id 3)}.

\subsection{Shallow Embeddings and OOP}
Shallow embeddings define a language directly through encoding its semantics
using procedural abstraction. In the case of \dsl,
an shallow embedded implementation should conform to the following
types:

@@

>>>>

\noindent The DSL consists of three constructs: two primitives
\emph{id} and \emph{fan} and one combinator \emph{beside}.
Their meanings are: \emph{id n} contains \emph{n} isolated vertical wires;
\emph{fan n} has \emph{n} vertical wires with its first wire connected to
all the remaining wires from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
With these three constructs, simple circuits can be described.
%For example, Figure~\ref{} visualize the circuit \emph{beside (fan 3) (id 3)}.

\subsection{Shallow Embeddings and OOP}\label{subsec:shallow}
Shallow embeddings define a language directly through encoding its semantics
using procedural abstraction. In the case of \dsl,
an shallow embedded implementation should conform to the following
types:

@@

Note that, for this simple circuit DSL, the Haskell domain is simply
\lstinline{Int}. This domain is a degenerate case of
procedural abstraction, where \lstinline{Int} can be viewed 
as a no argument function. In Haskell, due to laziness, \lstinline{Int}
is a good representation. In a call-by-value language 
a no-argument function \lstinline{() -> Int} would be more
appropriate to deal correctly with potential control-flow 
language constructs. More realistic shallow DSLs, such as parser 
combinators~\cite{leijen01parsec}, tend to have more complex functional domains.
